\LARGE \bf Artificial Cognition for Social Human-Robot Interaction:\\ An Implementation

### 2016-07-04 paper.tex
Second round of proof-reading 3

Humans commonly refer to the positions of objects with symbolic descriptors
(like \emph{on}, \emph{next to}...) instead of precise, absolute positions
(qualitative spatial reasoning). These type of descriptors have been studied in the context of language grounding
\cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.  {\sc
Spark} distinguishes between agent-independent symbolic locations (allocentric
spatial relations) and agent-dependent, relative locations (egocentric spatial
relations).

{\sc Spark} computes three main agent-independent relations~\cite{Sisbot2011}
based on the bounding box and centre of mass of the objects
(Figure~\ref{fig|spark-oro}\emph{a}): \concept{isOn} holds when an object $O_1$
is on another object $O_2$, and is computed by evaluating the centre of mass of
$O_1$ according to the bounding box of $O_2$.  \concept{isIn} evaluates if an
object $O_1$ is inside another object $O_2$ based on their bounding boxes
$BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
is next to another object $O_2$. We do not use a simple distance threshold to
determine if two objects are next to each other since the relation is highly
dependent on the dimensions of the objects. For instance, the maximum distance
between large objects (\eg two houses) to consider them as being next to each
other is much larger than the maximum distance we would consider for two small
objects (\eg two bottles). Thus, the distance threshold is scaled with the
objects' size. {\sc Spark} also compute symbolic facts related to agent
independent world dynamics.  The predicate \concept{isMoving} states, for each
tracked entity, whether it is currently moving or not.

Many other topological relations are dependent from the observation point
(egocentric perspective).  The predicate \concept{hasRelativePosition}
represents such spatial relations between agents and objects that are agent
dependent.  We compute these spatial locations by dividing the space around the
referent (an agent) into $n$ regions based on arbitrary angle values relative to
the referent orientation (Figure~\ref{fig|spark-oro}\emph{b}).  For example, for
$n = 4$ we would have the space divided into \emph{front, left, right} and
\emph{back}. Additionally, two proximity values, \emph{near} and \emph{far}, are
also considered. The number of regions and proximity values can be chosen
depending on the context where the interaction takes place.

Through perspective taking, {\sc Spark} computes for each agent a symbolic
description of the relative positioning of objects in the environment.
Table~\ref{facts} summarises all the symbolic spatial relations computed by {\sc
Spark}.

@@

\item \emph{Sees}: this relation describes what the agent can see, \ie what is
    within its field of view (FOV). In our current implementation, this
    affordance is computed by dynamically placing an OpenGL camera at the
    location of the eyes and running occlusion checks from it.  In
    Figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
    illustrated with a grey cone (the wider one). While he is able to see the two
    small boxes on the table in front of him, the big box on his right is out of
    his FOV, and therefore, he is not able to see it.

@@

To ensure a sufficiently stable recognition, these three capabilities are post-filtered
with an hysteresis function at the geometric level.

\begin{itemize}
\item \emph{Reachable} enables the robot to estimate the agent's ability to
    reach for an object, which is instrumental for effective social task
    planning. For example, if the human asks the robot to give her an object,
    the robot must compute a transfer point where she is able to get the
    object afterwards.  Reachability is computed for each agent (human or robot) based on
    Generalised Inverse Kinematics and collision detection. More than that, our robot is able 
    to compute an estimate of the effort needed by an agent to reach an object~\cite{pandey2013affordance}.

@@

Monitoring human activity is needed by the execution controllers to track the
engagement of the human and the progress of her actions and also to synchronise seamlessly
its own actions with the human actions. Full human action
and activity recognition is a task that requires knowledge and reasoning both on
high-level facts like goals, intentions and plans, as well as bottom-up data
from human and object motions. {\sc Spark} implements a set of simple temporal and
geometric heuristics on human hand trajectories and possible objects placements
to recognise simple elementary actions. Those primitive actions are assessed through
monitoring situations like ``an empty hand is close to an object on a table''
(precursor for a \emph{pick}), or ``a hand holding an object is over a
container'' (precursor for a \emph{throw}).  {\sc Spark} recognises a set of
such primitives. When combined with the other geometric computations and a predictive
plan of the human actions (see Section~\ref{hatp}), the execution controller can
track the fulfilment of the pre- and post-conditions of the predicted human
actions. As a result, the robot can monitor the engagement of the human and the
overall progress of the human-robot shared plan.

@@

In its current form, our situation assessment module makes two strong
assumptions: the objects are known in advance (hence, we can rely on proper 3D
CAD model for spatial reasoning) and the robot benefits of an almost perfect
perception, made possible by the use of fiducial markers.  Each object receives
a unique tag which enables us to accurately localise it in 3D and prevent
recognition ambiguities that would be then reflected in the knowledge base.
While {\sc Spark} algorithms themselves are ignorant of the input sources, and
would work equally well with a full object recognition stack, we are yet to
investigate this research area.

Additionally, temporal reasoning (essential for accurate action recognition for
instance) is not properly addressed in the current version of our system. Temporal reasoning is used
only locally, and does not allow for tracking of long sequences or global events.

@@

Natural language is a basic interaction modality that we use in our system both
as an input (processing of the human speech) and as an output (verbalization of the
robot intentions as well human-robot shared plans). Natural language processing is facilitated by our architecture
that processes information with semantics that are already close to a
human-level. This section presents the main features of our speech processor,
{\sc Dialogs}, that include semantic and multi-modal grounding, and interactive
disambiguation. Algorithmic and implementation details are provided
in~\cite{Lemaignan2011a}.

We acquire natural speech input from the human participants through a custom
Android-based interface. The interface relies on the Google speech recognition API for
speech-to-text and relays the textual transcript to the robot. The text is parsed into
a grammatical structure (\emph{Part of Speech} tagging), and the atoms of each
sentence are resolved with the help of the ontology to ground concepts like
objects (\ie when a user says ``pick up the can'', it resolves to which instance of
\emph{can} the user is referring to) and actions.  Figure~\ref{dialogs|ex} gives
an example of the processing of a simple, non-ambiguous command. The first
study (Section~\ref{moving-london}) walks through more complex examples.
Heuristics, like the presence of a question mark or the use of imperative mood,
are used to classify the sentences into questions, desires or statements.
{\sc Dialogs} processes these accordingly by answering questions or
updating the knowledge base.

\begin{figure}
    \centering
    \begin{tabular}{l}
        \emph{Initial human knowledge} \\
        \hline
        \stmt{BOOK1 type Book} \\
        \stmt{HUMAN1 type Human} \\
        ~\\
        ~\\
        ~\\
    \end{tabular}
    \begin{tabular}{l}
        \emph{Input}\\

@@

        \emph{Newly created statements}\\
        \hline
        \stmt{HUMAN1 desires SIT1} \\
         \stmt{SIT1 type Give} \\
         \stmt{SIT1 performedBy myself} \\
         \stmt{SIT1 actsOnObject BOOK1} \\
         \stmt{SIT1 receivedBy HUMAN1} \\
    \end{tabular}

@@

The system supports quantification (``give me \{a | the | some | all | any |
$n$\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), and anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also supports knowledge extension by learning new semantic
structures. For instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal} and added to the knowledge base
after checking for possible contradictions with existing knowledge. {\sc
Dialogs} finally interprets common temporal and spatial adverbs (like
\emph{above} or \emph{tomorrow}) and translates simple expressions of internal
state  into \emph{experiences} (for instance, ``I'm tired'' is processed into
\stmt{HUMAN1 experiences STATE1, STATE1 hasFeature tired}, see also
Section~\ref{sect:desires}). A full account of the {\sc Dialogs} features and
the corresponding algorithms is available in~\cite{Lemaignan2011a}.

\subsubsection{Dealing with Multi-Modality}

Because the components of our architecture rely on the same RDF formalism to
represent their outputs, the different communication modalities are presented in
a homogeneous way, as symbolic statements in the knowledge base. This applies
both to \emph{explicit} modalities (verbal communication, deictic gestures,
social gaze), and \emph{implicit} modalities (like the body position of the
human). The dialogue grounding process makes use of them at two distinct levels
to provide multi-modal concept grounding.

@@

Note that while the system benefits from complementary modalities, they are not
all required. The system can run with the verbal modality alone, at the cost of
a simpler interaction. For example, if the human says ``this'' without the robot
tracking what the human points at, no \stmt{HUMAN1 pointsAt ...} fact would
possibly be available in the knowledge base, and the robot falls back on the
anaphora resolution step alone.

The second level of integration of multi-modality is implicit. By continuously
computing symbolic properties from the geometric model, richer symbolic
descriptions are available to the system to verbalise or discriminate entities.
For instance, the robot may compute that one bottle is next to a glass, while
another one stands alone. These symbolic descriptions are transparently
re-used in a dialogue context to generate unambiguous references to discriminate
between similar objects: ``do you mean the bottle that is next to the glass?''.
The physical context of the interaction is used as an implicit communication
modality by {\sc Dialogs}. \cite{Ros2010b} provides a detailed account of our
approach toward interactive concept clarification and discrimination, along with the
related algorithms.

@@

The HATP planning framework extends the traditional Hierarchical Task
Network (HTN) planning domain representation and semantics by making them more
suitable to produce plans which involve humans and robots acting together
toward a joint goal. HATP is used by the robot to produce human-robot
\emph{shared plans}~\cite{Grosz1996,Clark1996,Kemp2007} which are then used to
anticipate human action, to suggest a course of action to humans, or even to ask
help from the human if needed.

@@

We briefly discuss here below how HATP incrementally builds and synchronises streams of actions concerning each or the agents (humans and robot) involved in a task, and how it promotes plans that satisfy humans needs and preferences as well as comfort and legibility issues.

@@

The originality of HATP resides in its ability to produce \emph{shared plans} which might involve 
the robot as well as the other participants (humans or robots).

@@

The planner also generates causal links and synchronisation points between the robot and his human partners. For
instance, in the plan depicted on Figure~\ref{plan_hatp1}, the human needs to
wait for the success of the robot's action {\tt PUTRV}. The robot monitors the
success of its own action (by checking for the fulfilment of the action
post-conditions; in this particular case \stmt{HUMAN sees BOTTLE} and
\stmt{HUMAN canReach BOTTLE}) to estimate what and when the human is likely to
perform his next action (here, \concept{TAKE(BOTTLE, TABLE)}). This, in turn,
allows the robot to monitor the human engagement and the progress in the shared
plan execution.

@@

A cost and a duration function are associated to each action.  The duration
function provides a duration interval for the action achievement and is used, in
one hand, to schedule the different streams and, on the other hand, as an
additional cost function. Cost here is not used here to estimate only standard time and energy consumption but
to estimate a number factors that measure the satisfaction of the human in terms
of acceptability and legibility of robot behaviour.

HATP includes mechanisms called \emph{social rules} to promote plans that are
considered as suitable for human-robot interaction. The planner
allows for the specification of the following filtering criteria:

\emph{Wasted time}: avoids plans where the human spends a lot
of her time being idle,

\emph{Effort balancing with respect of human desires}: avoids plans where efforts are
not distributed among the agents taking part to the plan following the human preferences.  It is
indeed sometimes beneficial to balance efforts between the human and
the robot. Sometimes the human wants to do more or, on the contrary, prefers to leave the 
robot do all what it can.

\emph{Simplicity}: promotes plans that limit as much as possible the interdependencies
between the actions of agents mentioned in the plan, as a problem with
executing just one of those actions could invalidate the entire
plan. Also intricate human-robot activity may cause discomfort since
the human will find himself repeatedly in a situation where his is
waiting for the robot to act,

@@

Combining the above criteria, we typically yield shared plans with desirable
interaction features like the human still being engaged in a number of tasks
while its overall level of required efforts remains low, or avoiding having the
human to wait for the robot (by essentially preventing the action streams from
having too many causal links between them).

Figure~\ref{plan_hatp2} illustrates such a socially-optimised plan where the
\emph{no wasted time} social rule is applied: compared to the plan depicted in
Figure~\ref{plan_hatp1}, the robot first moves the bottle so that the human can
immediately take it and put it into the bin, thus reducing the human idle time.
The resulting shared plan is considered better than the one depicted on
Figure~\ref{plan_hatp1} according to a global evaluation of the costs and rules.

In the current implementation, the social rules are effectively
implemented as filters and that are applied whenever a plan is produced.
In the future we intend to extend the plan-search algorithm and integrate the social rules 
in the process itself.

@@

As HATP is a generic symbolic task planner and does not enforce any abstraction
level for the planning domain, we were able to design a domain built from
top-level tasks whose semantics are close to the one used in the human-robot
dialogue: the planner domain effectively contains concepts like \texttt{give},
\texttt{table}, \texttt{is on}. This leads to an effective mapping between the
knowledge extracted from the situation assessment or the dialogue, and the
planner.

Also, we would like to mention that, after some research we have decided against
representing the planning domain (\ie, the set of tasks with their pre- and
post-conditions) in the knowledge base itself due to expressiveness issues (see
appendix B of~\cite{Lemaignan2012a} for a detailed discussion). This effectively
leads to independent declarative and procedural knowledge stores.

@@

While parts of the architecture ({\sc Spark}, {\sc Oro}) have been deployed with
external execution controllers (like {\sc Cram}~\cite{Beetz2010} or the BERT
platform~\cite{Lallee2010b}, as reported in~\cite{Lemaignan2010}), we have also
developed dedicated robot controllers which integrate the whole stack introduced
in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2009} is the main one,
written in the \emph{Procedural Reasoning System (PRS)} language~\cite{Ingrand1996}.
We have also developed the Python-based {\sc pyRobots}~\cite{lemaignan2015pyrobots} that
provides a large set of high-level actions and an event-based architecture well
suited for prototyping. They both rely on extensive integration with the knowledge
base, that serves as the primary source of semantics for the decision making process.

One of the main roles of {\sc Shary} is to control the production and execution
of shared plans. This means essentially context-based refinement and robot
actions execution, as well as monitoring of those achieved by its human
partner. One of the key design goals is to build such abilities in a generic
way, and to provide several levels of parametrisation allowing to adapt to
various environments, and various levels of involvement of the robot ranging
from teammate behaviour to assistant or proactive helper. Based on this,
the robot controller invokes the adequate human-aware planners and react to
events triggered by the {\sc Oro} knowledge base (as described below).

Execution control systems for social robots are expected 
to account not only for the task achievement but also for the communication
and monitoring acts needed to support human-robot interactive task
achievement~\cite{Rich1997,Sidner2005,sebanz2006joint}. {\sc Shary} supports
such mechanisms~\cite{Clodic2007smc}. It allows
to bind action execution to \emph{communication policies} in order to produce
multi-modal signals towards the human and to react to human actions and signals. The
\emph{communication act} is the central concept in this formalism. It
represents an information exchange between the two agents and plays the role of
a transition condition. This exchange can be realised through speech, through an
expressive motion or a combination of the two. It enables the robot as well as the human to
communicate their beliefs about the task to be achieved, in order to share
mutual knowledge and to synchronise their activities.  This is done through
real-time task-based situation assessment achieved by the combination of {\sc
Shary} monitoring processes and {\sc Oro} inference mechanisms. In {\sc Shary},
each action description contains not only how to execute it and monitor it,
but also how to initiate it (\eg in the case of a shared action, how
both agents need to commit to the task) and how to suspend (or cancel) it. In
each of these cases, the action description also makes explicit how the
transition has to be communicated to the human partner if necessary. This is
supported by a description of the beliefs of each of the agents
regarding the action achievement process, as well as a set of so-called
\emph{monitoring processes} that update them.

@@

The {\sc Oro} server supports two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. A module can
subscribe to an event by passing through an event pattern (in its simplest
form, a partial statement like \stmt{? type Book}) and a callback.  Each
time a new instance of a book appears in the knowledge base, the callback is
triggered.

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{HUMAN1 desires
?action, ?action type Give, ?action actsOnObjecT ?obj, ?obj type Book}, we could
trigger a behaviour when the human expresses (through dialogue, gestures...)
that he wants the robot to give her a book.

The robot controller designer does not need to directly care about how this
\emph{desire} is produced (this is delegated to perception modules), he can
focus on the semantics of the desire.

Note also that we take advantage of the reasoning capabilities of the system:
for example, the type of the object (\stmt{?obj type Book}) may not be
explicitly asserted, but inferred by the reasoner based on other assertions.

@@

\emph{Desires} are typically human commands (``Give me that book''). The nature of
the desired action (to pick, to give, to look, to bring, to show...), along with the action
parametrization (thematic roles) are extracted from the knowledge base, and
either passed to the task planner or executed if the procedure is directly
available.

@@

\subsubsection{Human-Aware Action Refinement, Execution and Monitoring}\label{sec:action}

Like most robotic architectures, actions are split into \emph{elementary actions}
that are combined into \emph{tasks}. Tasks are created either statically or dynamically.
The {\sc pyRobots} controller, for instance, statically combines the actions listed in
table~\ref{table|pyrobots_actions} to implement the high-level tasks it exposes: {\tt
lookAt}, {\tt moveTo}, {\tt getFromHuman}, {\tt showObject}, {\tt giveObject},
{\tt pickObject}, {\tt bringObject}, {\tt putObject}, {\tt hideObject}.
Our other controller, {\sc Shary}, relies on the symbolic task planner to
dynamically generate, at run-time, suitable sequence of elementary actions.

>>>>

Humans commonly refer to the positions of objects with symbolic descriptors
(like \emph{on}, \emph{next to}...) instead of precise, absolute positions
(qualitative spatial reasoning). These type of descriptors have been extensively studied in the context of language grounding
\cite{O'Keefe1999,Matuszek2010,Regier2001,Kelleher2006,Blisard2005}.  {\sc
Spark} distinguishes between agent-independent symbolic locations (allocentric
spatial relations) and agent-dependent, relative locations (egocentric spatial
relations).

{\sc Spark} computes three main agent-independent relations
based on the bounding box and centre of mass of the objects
(Figure~\ref{fig|spark-oro}\emph{a},~\cite{Sisbot2011}): \concept{isOn} holds when an object $O_1$
is on another object $O_2$, and is computed by evaluating the centre of mass of
$O_1$ according to the bounding box of $O_2$.  \concept{isIn} evaluates if an
object $O_1$ is inside another object $O_2$ based on their bounding boxes
$BB_{O_1}$ and $BB_{O_2}$.  \concept{isNextTo} indicates whether an object $O_1$
is next to another object $O_2$. Note that we do not use a simple distance threshold to
determine if two objects are next to each other since the relation is highly
dependent on the dimensions of the objects. For instance, the maximum distance
between large objects (\eg two houses) to consider them as being next to each
other is much larger than the maximum distance we would consider for two small
objects (\eg two bottles). Thus, the distance threshold is scaled with the
objects' size. Finally, {\sc Spark} also computes symbolic facts related to agent
independent world dynamics.  The predicate \concept{isMoving} states, for each
tracked entity, whether it is currently moving or not.

Many other topological relations are dependent from the observation point
(egocentric perspective).  The predicate \concept{hasRelativePosition}
represents the superset of such spatial relations between agents and objects
that are agent-dependent. We compute these spatial relations by dividing the space around the
referent (an agent) into $n$ regions based on arbitrary angle values relative to
the referent orientation (Figure~\ref{fig|spark-oro}\emph{b}).  For example, for
$n = 4$ we would have the space divided into \emph{front, left, right} and
\emph{back}. Additionally, two proximity values, \emph{near} and \emph{far}, are
also considered. The number of regions and proximity values can be chosen
depending on the context where the interaction takes place.
Through perspective taking, {\sc Spark} computes for each agent a symbolic
description of the relative positioning of objects in the environment.

Table~\ref{facts} summarises all the symbolic spatial relations computed by {\sc
Spark}.

@@

\item \emph{Sees}: this relation describes what the agent can see, \ie what is
    within its field of view (FOV). In our current implementation, this
    affordance is computed by dynamically placing an OpenGL camera at the
    location of the eyes and running occlusion checks from it.  In
    Figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
    illustrated with a grey cone (the wider one). While she is able to see the two
    small boxes on the table in front of her, the big box on her right is out of
    her FOV, and therefore, she is not able to see it.

@@

We apply an hysteresis filter at the geometric level to ensure a sufficiently
stable recognition of these three capabilities.

\begin{itemize}
\item \emph{Reachable} enables the robot to estimate the agent's ability to
    reach for an object, which is instrumental for effective social task
    planning. For example, if the human asks the robot to give her an object,
    the robot must compute a transfer point where she is able to get the
    object afterwards.  Reachability is computed for each agent (human or robot) based on
    Generalised Inverse Kinematics and collision detection. Besides, the robot is able 
    to compute an estimate of the effort needed by an agent to reach an object~\cite{pandey2013affordance}.

@@

Monitoring human activity is needed by the execution controllers to track the
engagement of the human and the progress of their actions. It is also needed to synchronise seamlessly
its own actions with the human actions. Full human action
and activity recognition is a task that requires knowledge and reasoning both on
high-level facts like goals, intentions and plans, as well as bottom-up data
from human and object motions. {\sc Spark} implements a set of simple temporal and
geometric heuristics on human hand trajectories and possible objects placements
to recognise simple elementary actions. Those primitive actions are assessed by
monitoring situations like ``an empty hand is close to an object on a table''
(precursor for a \emph{pick}), or ``a hand holding an object is over a
container'' (precursor for a \emph{put}).  {\sc Spark} recognises a set of
such primitives. When combined with the other geometric computations and a predictive
plan of the human actions (see Section~\ref{hatp}), the execution controller can
track the fulfilment of the pre- and post-conditions of the predicted human
actions. The robot relies on these to monitor the engagement of the human and the
overall progress of the human-robot shared plan.

@@

In its current form, our situation assessment module makes two 
assumptions: the objects are known in advance (hence, we can rely on proper 3D
CAD model for spatial reasoning) and the robot benefits of an nearly perfect
perception, made possible by the use of fiducial markers.  Each object receives
a unique tag which enables an accurate localisation in 3D and prevents
recognition ambiguities that would be otherwise reflected in the knowledge base.
While {\sc Spark} algorithms do not concern themselves with the nature of the input sources, and
would work equally well with a full object recognition stack, we did not
investigate this research area so far.

Additionally, temporal reasoning (essential for accurate action recognition for
instance) is not generally addressed in the current state of our system. Temporal reasoning is used
only locally, and does not allow for tracking of long sequences or global events.

@@

Natural language is a basic interaction modality that we use in our system both
as an input (processing of the human speech) and as an output (verbalization of the
robot intentions, as well as human-robot shared plans). Natural language
processing is facilitated as our architecture
manipulates semantics that are close to the
human level. This section presents the main features of our speech processor,
{\sc Dialogs}, that include semantic and multi-modal grounding, and interactive
disambiguation. Algorithms and implementation details are provided
in~\cite{Lemaignan2011a}.

We acquire natural speech input from the human participants through a custom
Android-based interface. The interface relies on the Google speech recognition API for
speech-to-text (ASR) and relays the textual transcript to the robot. The text is parsed into
a grammatical structure (\emph{Part of Speech} tagging) by a custom
heuristics-based parser. The resulting atoms are then resolved with the help of
the knowledge base to ground concepts like
objects (\ie when a user says ``pick up the can'', it resolves to which instance of
\concept{Can} the user is referring to) and actions.  Figure~\ref{dialogs|ex} gives
an example of the processing of a simple, non-ambiguous command. The first
study (Section~\ref{moving-london}) walks through more complex examples.
Heuristics, like the presence of a question mark or the use of imperative mood,
are used to classify the sentences into questions, desires or statements.
{\sc Dialogs} processes these accordingly by answering questions or
updating the knowledge base.

\begin{figure}
    \centering
    \begin{tabular}{l}
        \emph{Initial human model} \\
        \hline
        \stmt{BOOK1 type Book} \\
        \stmt{HUMAN1 type Human} \\
        ~\\
        ~\\
        ~\\
    \end{tabular}
    \begin{tabular}{l}
        \emph{Input}\\

@@

        \emph{Statements added to robot model}\\
        \hline
        \stmt{HUMAN1 desires SITUATION1} \\
         \stmt{SITUATION1 type Give} \\
         \stmt{SITUATION1 performedBy myself} \\
         \stmt{SITUATION1 actsOnObject BOOK1} \\
         \stmt{SITUATION1 receivedBy HUMAN1} \\
    \end{tabular}

@@

The system supports quantification (``give me \{a | the | some | all | any |
$n$\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), and anaphora resolution (``give \emph{it} to me'') based on
the dialogue history and the working memory. It also supports knowledge extension by learning new semantic
structures. For instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal} and added to the knowledge base
after checking for possible contradictions with existing knowledge. {\sc
Dialogs} finally interprets common temporal and spatial adverbs (like
\emph{above} or \emph{tomorrow}) and translates simple expressions of internal
state  into \emph{experiences} (for instance, ``I'm tired'' is processed into
\stmt{HUMAN1 experiences STATE1, STATE1 hasFeature tired}, see also
Section~\ref{sect:desires}). A full account of the {\sc Dialogs} features and
the corresponding algorithms is available in~\cite{Lemaignan2011a}.

\subsubsection{Dialogue and Multi-Modality}

Because all of the components of our architecture rely on the same RDF formalism to
represent their outputs, the different communication modalities are presented in
a homogeneous way, as symbolic statements in the knowledge base. This applies
both to \emph{explicit} modalities (verbal communication, deictic gestures,
gaze focus), and \emph{implicit} modalities (like the body position of the
human). The dialogue grounding process makes use of them at two distinct levels
to provide multi-modal concept grounding.

@@

Note that while the system benefits from complementary modalities, they are not
all required. The system can run with the verbal modality alone, at the cost of
a simpler interaction. For example, if the human says ``this'' without the robot
tracking what the human points at, no \stmt{HUMAN1 pointsAt ...} fact is
possibly available in the knowledge base, and the robot falls back on the
anaphora resolution step alone.

The second level of integration of multi-modality is implicit. By continuously
computing symbolic properties from the geometric model, richer symbolic
descriptions are available to the system to verbalise or discriminate entities.
For instance, the robot may compute that one bottle is next to a glass, while
another one stands alone. These symbolic descriptions are transparently
re-used in a dialogue context to generate unambiguous references to discriminate
between similar objects: ``do you mean the bottle that is next to the glass?''.
The physical context of the interaction is used as an implicit communication
modality by {\sc Dialogs}. \cite{Ros2010b} provides a detailed account of our
approach to interactive concept clarification and discrimination, along with the
related algorithms.

@@

The HATP planning framework extends the traditional Hierarchical Task
Network (HTN) planning domain representation and semantics by making them more
suitable to produce plans which involve humans and robots acting together
toward a joint goal. HATP is used by the robot to produce human-robot
\emph{shared plans}~\cite{Grosz1996,Clark1996,Kemp2007} which are then used to
anticipate human action, to suggest a course of action to humans, or possibly to ask
help from the human if needed.

@@

We discuss next how HATP incrementally builds and synchronises streams of
actions for each or the agents (humans and robot) involved in a task, and how it
promotes plans that satisfy humans needs and preferences as well as comfort and
legibility.

@@

The originality of HATP resides in its ability to produce \emph{shared plans} which might involve 
the robot as well as the other participants.

@@

The planner also generates causal links and synchronisation points between the robot and his human partners. For
instance, in the plan depicted on Figure~\ref{plan_hatp1}, the human needs to
wait for the success of the robot's action {\tt PUTRV}. The robot monitors the
success of its own action (by checking for the fulfilment of the action
post-conditions; in this particular case \stmt{HUMAN sees BOTTLE} and
\stmt{HUMAN canReach BOTTLE}) to estimate what and when the human is likely to
perform his next action (here, \concept{TAKE(BOTTLE, TABLE)}). This, in turn,
allows the robot to monitor the human engagement and the progress in the shared
plan execution. A complete example is presented in
Section~\ref{sec:cleaning-table}.

@@

A duration and a cost function are associated to each action. The duration
function provides a duration interval for the action achievement. It serves both
to schedule the different streams and as an additional cost function. Besides
time and energy consumption, the cost function integrates factors that measure
the satisfaction of the human in terms of acceptability and legibility of the
resulting robot behaviour.

HATP includes mechanisms called \emph{social rules} to promote plans that are
considered as suitable for human-robot interaction. The following constraints
can be set:

\emph{Wasted time}: avoids plans where the human spends a lot
of her time being idle;

\emph{Effort balancing with respect of human desires}: avoids plans where efforts are
not distributed among the agents taking part to the plan, respective of the human preferences.  It is
indeed sometimes beneficial to balance efforts between the human and
the robot. Sometimes the human wants to do more, or on the contrary, prefers to leave the 
robot do most of the work;

\emph{Simplicity}: promotes plans that limit as much as possible the interdependencies
between the actions of agents involved in the plan, as an issue during the
execution of one of those actions would put the entire plan at risk. Also intricate human-robot activity may cause discomfort since
the human will find herself repeatedly in a situation where she is
waiting for the robot to act;

@@

Combining the above criteria, we yield shared plans with desirable
interaction features like having the human engaged in a number of tasks
while her overall level of efforts remains low, or avoiding having the
human to wait for the robot by preventing the action streams from
having too many causal links between them.

Figure~\ref{plan_hatp2} illustrates such a socially-optimised plan where the
\emph{no wasted time} social rule is applied: compared to the plan depicted in
Figure~\ref{plan_hatp1}, the robot first moves the bottle so that the human can
immediately take it and put it into the bin, thus reducing the human idle time.

In the current implementation, the social rules are effectively implemented as
filters applied to the set of all possible plans computed by the planner.  In
the future, we intend to extend the plan-search algorithm and integrate the
social rules in the process itself.

@@

As HATP is a generic symbolic task planner and does not enforce any abstraction
level for the planning domain, we have designed a planning domain made of
top-level tasks whose semantics are close to the one used in the human-robot
dialogue: the planner domain effectively contains concepts like \texttt{Give},
\texttt{table}, \texttt{isOn}. This leads to an effective mapping between the
knowledge extracted from the situation assessment or the dialogue, and the
planner.

Due to expressiveness issues, we do not represent the planning domain (\ie, the set of tasks with their pre- and
post-conditions) in the knowledge base directly (see
appendix B of~\cite{Lemaignan2012a} for the full rationale). This effectively
leads to independent declarative (the {\sc Oro} knowledge base) and procedural
(the planner) knowledge stores.

@@

While parts of the architecture ({\sc Spark}, {\sc Oro}) have been deployed with
external execution controllers (like {\sc Cram}~\cite{Beetz2010} or the BERT
platform~\cite{Lallee2010b}, as reported in~\cite{Lemaignan2010}), we have also
developed dedicated robot controllers which integrate into the deliberative
architecture introduced
in Figure~\ref{fig|archi}. {\sc Shary}~\cite{clodic2009} is the main one,
written in the \emph{Procedural Reasoning System (PRS)} language~\cite{Ingrand1996}.
We have also developed the Python-based {\sc pyRobots}~\cite{lemaignan2015pyrobots} that
provides a large set of high-level actions and an event-based architecture well
suited for prototyping. They both rely on extensive integration with the knowledge
base, that serves as the primary source of semantics for the decision making process.

One of the main roles of {\sc Shary} is to control the production and execution
of shared plans. This means essentially context-based refinement and execution
of actions, as well as monitoring of those achieved by its human
partner. One of the key design goals is to build such abilities in a generic
way, and to provide several levels of parametrisation allowing to adapt to
various environments and various levels of involvement of the robot, ranging
from teammate behaviour to assistant or proactive helper. Based on this,
the robot controller invokes the adequate human-aware planners and react to
events triggered by the {\sc Oro} knowledge base (as described below).

Execution control systems for social robots are expected 
to account not only for the task achievement but also for the communication
and monitoring acts needed to support human-robot interactive task
achievement~\cite{Rich1997,Sidner2005,sebanz2006joint}. {\sc Shary} supports as
well
such a mechanism~\cite{Clodic2007smc}. It allows
to bind action execution to \emph{communication policies} in order to produce
multi-modal signals towards the human and to react to human actions and signals. The
\emph{communication act} is the central concept in this formalism. It
represents an information exchange between the two agents and plays the role of
a transition condition. This exchange can be realised through speech, through an
expressive motion or a combination of the two. It enables the robot as well as the human to
communicate their beliefs about the task to be achieved, in order to share
mutual knowledge and to synchronise their activities.  This is done through
real-time task-based situation assessment achieved by the combination of {\sc
Shary} monitoring processes and {\sc Oro} inference mechanisms. In {\sc Shary},
each action description contains not only how to execute it and monitor it,
but also how to initiate it (\eg in the case of a shared action, how
both agents need to commit to the task) and how to suspend (or cancel) it. In
each of these cases, the action description also makes explicit how the
transition has to be communicated to the human partner if necessary. This is
supported by a description of the beliefs of each of the agents
regarding the action achievement process, as well as a set of so-called
\emph{monitoring processes} that update them.

@@

The {\sc Oro} server supports two paradigms to access its content: RPC-style
queries (based on the standard SPARQL language) or events. A module can
subscribe to an event by registering an event pattern (in its simplest
form, a partial statement like \stmt{? type Book}) and a callback.  Each
time a new instance of a book appears in the knowledge base, the callback is
triggered.

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{HUMAN1 desires
?action, ?action type Give, ?action actsOnObject ?obj, ?obj type Book}, we
effectively trigger a behaviour when the human expresses (through dialogue,
gestures...) that he wants the robot to give her a book.

The robot controller designer does not need to directly care about how this
\emph{desire} is produced (this is delegated to perception modules), he can
instead focus on the semantics of the desire.

Note that we transparently take advantage of the reasoning capabilities of the
system to trigger event to trigger events:
for example, the type of the object (\stmt{?obj type Book}) may not be
explicitly asserted, but inferred by the reasoner based on other assertions.

@@

\emph{Desires} are typically human commands (``Give me that book''). The nature of
the desired action (to pick, to give, to look, to bring, to show...), along with the action
parametrization (thematic roles) are fetched by the execution controller in the knowledge base, and
are either sent as goals to the task planner, or executed if the elementary
action is directly available.

@@

\subsubsection{Human-Aware Action Refinement}\label{sec:action}
